Memory-Efficient Mixed-Precision Implementations\\for Robust Explicit Model Predictive Control

### 2019-04-11 emsoft-2019/RMPC_Daisy.tex
two more comments
comments and edits for section 3

% !TEX root = main.tex
\section{Explicit MPC and Finite Precision}
\subsection{Robust Explicit MPC}
\def\reals{\mathbb{R}}
We consider the class of linear time-invariant (LTI) systems characterized by the difference equation
\begin{equation}
\label{eq:DSS}
x_{k+1}=Ax_k+Bu_k+Ew_k,\quad  k=0,1,2,\ldots
\end{equation}
where $x\in \reals^{n\times 1}$ is the state, $u\in \reals^{m\times 1}$ is the control input and $w\in \reals^{d\times 1}$ is the disturbance. Matrices $A\in \reals^{n\times n}$, $B\in \reals^{n\times m}$ and $E\in \reals^{n\times d}$ are capturing respectively the effects of current state, input and disturbance on the next state. We assume the disturbance $w$ belongs to a set $\mathcal{W}$ where $\mathcal{W}$ is a polyhedral set.
In this paper, we focus on robust formulation of MPC which at each time step, minimizes the worst-case value of the objective function with respect to the disturbances over the control inputs. We also consider the objective function to be quadratic with respect to the states and inputs. To reduce the the conservativeness of the optimization and enlarge the set of feasible input trajectories, we assume the pair $(A,B)$ is stablizable and
$u_k=Kx_k+v_k$, where $K$ is selected such that some desired property such as stability is satisfied, and synthesize $v_k$ for the system.
%The modifed state evolution of the system will be $x_{k+1}=\bar Ax_k+Bv_k+Ew_k$ with $\bar A := A +BK$.
% and $v_k=\kappa(x_k)$ denotes the output of model predictive controller.  
The constrained optimization at each time step is of the form
\begin{align}
\label{eq:RMPC_prob}
J^{\ast}(x_0)=\min_{v_0,\cdots,v_{N-1}}& \max_{w_0,\cdots,w_{N-1}} \sum_{i=0}^{N-1}(x_i^TQx_i+u_i^TRu_i) + x_N^TQ_Fx_N\nonumber\\
\text{s.t.} \quad &x_{i+1}=Ax_i+Bu_i + E w_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
\quad &u_i=Kx_i+v_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
&u_i\in\mathcal{U},\quad \forall w_i\in\mathcal{W}, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
&x_i\in\mathcal{X},\quad \forall w_i\in\mathcal{W},\quad\forall i\in\{0,1,\cdots,N\},
\end{align}
%\Sadegh{The role of disturbance is missing in the inequalities. We should say, we want to satisfy the inequalities for all possible values of the disturbance.}
%
where $\mathcal U$ and $\mathcal X$ are polyhedral sets as feasible sets of inputs and states. Matrices $Q\in\reals^{n\times n}$ and $Q_F\in\reals^{n\times n}$ are positive definite, $R\in\reals^{m\times m}$ is positive semidefinite and $N$ denotes the length of the prediction horizon.
\begin{theorem}[\cite{delaPea:2005}]
The optimization  \eqref{eq:RMPC_prob} can be translated into a multi-parametric quadratic problem which admits a closed-form solution. Furthermore, for the case that $R>0$, the controller is $u_k = K x_k + \kappa(x_k)$ with $\kappa(\cdot)$ being a continuous piecewise affine (PWA) function over polyhedral regions:
\begin{equation}
\kappa(x_k)=
\begin{cases}
F_1x_k+G_1 & \text{if $x_k\in \mathcal{R}_1$}\\
F_2x_k+G_2 & \text{if $x_k\in \mathcal{R}_2$}\\
\vdots\\
F_Px_k+G_P & \text{if $x_k\in \mathcal{R}_P$}
\end{cases} 
\end{equation}
where $\mathcal{R}_i$ is a ploytope determined by a set of linear inequalities $\mathcal R_i = \{x\in\mathcal X\,|\,H_ix\leq K_i\}$. 
\end{theorem}
The essential idea behind this theorem is to utilize the closed form $x_i=A^ix_0+\sum_{l=0}^{i-1}A^{i-l-1}Bu_l$ and transform the optimization \eqref{eq:RMPC_prob} into the following quadratic program:
%\begin{align}
%	&\min_{U_N}U_N^TZU_N+x_0^TVU_N+x_0^TYx_0\nonumber\\
%	&s.t.\quad JU_N\leq s+Dx_0
%\end{align} 
\begin{align}
\label{eq:multi_param_prog}
J^{\ast}(x_0)=\min_{z,\gamma}& \left[x_0^TYx_0+\frac{1}{2}z^THz+\gamma\right]\\
\text{s.t.} \quad &G_mz+g_m\gamma\leq W_m+S_mx_0,\nonumber\\
&G_cz\leq W_c+S_cx_0\nonumber
\end{align}
where, $z\in \reals^{mN}$ and $\gamma\in\reals$ are decision variables and $Y$, $H$, $G_m$, $G_c$,$S_m$, $S_c$, $W_m$, $W_c$, $g_m$ are matrices of proper sizes that can be easily obtained from the original optimization \eqref{eq:RMPC_prob}  (see \cite{delaPea:2005}).

@@

The optimization \eqref{eq:multi_param_prog} can be solved using multi-parametric techniques to compute the explicit form of $\kappa(\cdot)$.
Efficient implementation of such computations is available in multi-parametric toolbox of MATLAB~\cite{matlabMPT, matlabYALMIP}.
Let us denote the set of states $x_0$ for which the optimization  \eqref{eq:multi_param_prog} is feasible by $\mathcal X_s$. Then $\mathcal X_s\subseteq \mathcal X$ and is is the union of all regions $\mathcal{R}_{i}$:
\begin{equation}
\mathcal X_s = \bigcup_{i}(H_{i}\statevar<=K_{i}).
\end{equation}

@@

%Therefore, given the state vector $x$, to compute control input, one has to first determine to which polytope does the state vector belong.
Several algorithms are proposed in the literature \cite{Mnnigmann:2011,Jones:2006} to find the right polytope $i$ to which the state $x_k$ belong. The most straightforward technique is a linear search over all polytopes in \statespace\space using a switch/case conditional statement of Listing \ref{lst:caseof}.
%\Sadegh{Perhaps we can mention another approach here? add references.}

\begin{lstlisting}[escapeinside={(*}{*)},label={lst:caseof}, caption=switch for region selection]
switch((*\statevarmath*)):
case (*$ H_{1}\statevar<=K_{1}$*) then (*$v_{k}=F_1x_k+G_1 $*)
case (*$ H_{2}\statevar<=K_{2}$*) then (*$v_{k}=F_2x_k+G_2$*)
case (*$ H_{3}\statevar<=K_{3}$*) then (*$v_{k}=F_3x_k+G_3$*)
...
case (*$ H_{P}\statevar<=K_{P}$*) then (*$v_{k}=F_Px_k+G_P$*)
\end{lstlisting}
%Once the polytope was determined, computing for $F_ix+G_i$ requires even less effort as it only involves summation and multiplication.

To keep the costs of implementation down, low-end microcontrollers usually have limited memory and computational power. Therefore, implementation of the above explicit MPC on microcontrollers will bring some issues since all regions' matrices ($H_i$s and $K_i$s) and activation functions ($F_i$s and $G_i$s) have to be stored on the microcontroller. The goal of our paper is present an approach for minimizing the memory usage of the implementation while satisfying the hard constraints on the states. 
%\section{Finite Precision Implementation}
%\Sadegh{To be included by Rocco}

>>>>

% !TEX root = main.tex
\section{Explicit MPC and Finite Precision}
In this section, we provide relevant background on robust explicit MPC and
fixed-point arithmetic and error analysis.

\subsection{Robust Explicit MPC}
\def\reals{\mathbb{R}}
We consider the class of linear time-invariant (LTI) systems characterized by the difference equation
\begin{equation}
\label{eq:DSS}
x_{k+1}=Ax_k+Bu_k+Ew_k,\quad  k=0,1,2,\ldots
\end{equation}
where $x\in \reals^{n\times 1}$ is the state, $u\in \reals^{m\times 1}$ is the control input and $w\in \reals^{d\times 1}$ is the disturbance. Matrices $A\in \reals^{n\times n}$, $B\in \reals^{n\times m}$ and $E\in \reals^{n\times d}$ are capturing respectively the effects of current state, input and disturbance on the next state. We assume the disturbance $w$ belongs to a set $\mathcal{W}$ where $\mathcal{W}$ is a polyhedral set.
In this paper, we focus on the robust formulation of MPC which at each time step
minimizes the worst-case value of the objective function\todo{which objective function? first mention here without explanation} with respect to the
disturbances over the control inputs. We consider the objective function to
be quadratic with respect to the states and inputs. To reduce the 
conservativeness of the optimization and enlarge the set of feasible input
trajectories, we assume the pair $(A,B)$ is stabilizable and
$u_k=Kx_k+v_k$, where $K$ is selected such that some desired property such as stability is satisfied, and synthesize $v_k$ for the system.
\eva{I don't understand the above sentence. Maybe split into several sentences? Where does $u_k$ come from (the sentence right now says you assume $u_k$)?}
%The modifed state evolution of the system will be $x_{k+1}=\bar Ax_k+Bv_k+Ew_k$ with $\bar A := A +BK$.
% and $v_k=\kappa(x_k)$ denotes the output of model predictive controller.  
The constrained optimization at each time step is of the form
\begin{align}
\label{eq:RMPC_prob}
J^{\ast}(x_0)=&\min_{v_0,\cdots,v_{N-1}} \max_{w_0,\cdots,w_{N-1}} \sum_{i=0}^{N-1}(x_i^TQx_i+u_i^TRu_i) + x_N^TQ_Fx_N\nonumber\\
\text{s.t.} \quad &x_{i+1}=Ax_i+Bu_i + E w_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
\quad &u_i=Kx_i+v_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
&u_i\in\mathcal{U},\quad \forall w_i\in\mathcal{W}, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
&x_i\in\mathcal{X},\quad \forall w_i\in\mathcal{W},\quad\forall i\in\{0,1,\cdots,N\},
\end{align}
%\Sadegh{The role of disturbance is missing in the inequalities. We should say, we want to satisfy the inequalities for all possible values of the disturbance.}
%
where $\mathcal U$ and $\mathcal X$ are polyhedral sets as\todo{of?} feasible
sets of inputs and states. Matrices $Q\in\reals^{n\times n}$ and
$Q_F\in\reals^{n\times n}$ are positive definite, $R\in\reals^{m\times m}$ is
positive semidefinite and $N$ denotes the length of the prediction horizon.
\eva{What are $Q, Q_F, R$ and where do they come from?}

\begin{theorem}[\cite{delaPea:2005}]
The optimization \eqref{eq:RMPC_prob} can be translated into a multi-parametric quadratic problem which admits a closed-form solution. Furthermore, for the case that $R>0$, the controller is $u_k = K x_k + \kappa(x_k)$ with $\kappa(\cdot)$ being a continuous piecewise affine (PWA) function over polyhedral regions:
\begin{equation}
\kappa(x_k)=
\begin{cases}
F_1x_k+G_1 & \text{if $x_k\in \mathcal{R}_1$}\\
F_2x_k+G_2 & \text{if $x_k\in \mathcal{R}_2$}\\
\vdots\\
F_Px_k+G_P & \text{if $x_k\in \mathcal{R}_P$}
\end{cases} 
\end{equation}
where $\mathcal{R}_i$ is a ploytope\todo{polytope?} determined by a set of linear inequalities $\mathcal R_i = \{x\in\mathcal X\,|\,H_ix\leq K_i\}$. 
\end{theorem}
The essential idea behind this theorem is to utilize the closed form $x_i=A^ix_0+\sum_{l=0}^{i-1}A^{i-l-1}Bu_l$ and transform the optimization \eqref{eq:RMPC_prob} into the following quadratic program:
%\begin{align}
%	&\min_{U_N}U_N^TZU_N+x_0^TVU_N+x_0^TYx_0\nonumber\\
%	&s.t.\quad JU_N\leq s+Dx_0
%\end{align} 
\begin{align}
\label{eq:multi_param_prog}
J^{\ast}(x_0)=\min_{z,\gamma}& \left[x_0^TYx_0+\frac{1}{2}z^THz+\gamma\right]\\
\text{s.t.} \quad &G_mz+g_m\gamma\leq W_m+S_mx_0,\nonumber\\
&G_cz\leq W_c+S_cx_0\nonumber
\end{align}
where, $z\in \reals^{mN}$ and $\gamma\in\reals$ are decision variables and $Y$, $H$, $G_m$, $G_c$,$S_m$, $S_c$, $W_m$, $W_c$, $g_m$ are matrices of proper sizes that can be easily obtained from the original optimization \eqref{eq:RMPC_prob}  (see \cite{delaPea:2005}).

@@

The optimization \eqref{eq:multi_param_prog} can be solved using multi-parametric techniques to compute the explicit form of $\kappa(\cdot)$.
An efficient implementation of such computations is available in the multi-parametric toolbox of MATLAB~\cite{matlabMPT, matlabYALMIP}.
Let us denote the set of states $x_0$ for which the optimization  \eqref{eq:multi_param_prog} is feasible by $\mathcal X_s$. Then $\mathcal X_s\subseteq \mathcal X$ and is the union of all regions $\mathcal{R}_{i}$:
\begin{equation}
\mathcal X_s = \bigcup_{i}(H_{i}\statevar<=K_{i}).
\end{equation}

@@

%Therefore, given the state vector $x$, to compute control input, one has to first determine to which polytope does the state vector belong.
Several algorithms are proposed in the literature \cite{Mnnigmann:2011,Jones:2006} to find the right polytope $i$ to which the state $x_k$ belong. The most straightforward technique is a linear search over all polytopes in \statespace\space using a switch/case conditional statement as shown in~\autoref{lst:caseof}.
%\Sadegh{Perhaps we can mention another approach here? add references.}

\begin{figure}[t]
\begin{lstlisting}[mathescape=true,basicstyle=\small\ttfamily,morekeywords={switch, case, then}]
switch($\statevar$):
  case ($H_{1}\statevar<=K_{1}$) then ($v_{k}=F_1x_k+G_1$)
  case ($H_{2}\statevar<=K_{2}$) then ($v_{k}=F_2x_k+G_2$)
  case ($H_{3}\statevar<=K_{3}$) then ($v_{k}=F_3x_k+G_3$)
    ...
  case ($ H_{P}\statevar<=K_{P}$) then ($v_{k}=F_Px_k+G_P$)
\end{lstlisting}
\caption{Structure of an explicit MPC controller}
\label{lst:caseof}
\end{figure}
%Once the polytope was determined, computing for $F_ix+G_i$ requires even less effort as it only involves summation and multiplication.

To keep the cost of implementation down, low-end microcontrollers usually have
limited memory and computational power. Therefore, implementation of the above
explicit MPC on microcontrollers will bring some issues since all regions'
matrices ($H_i$s and $K_i$s) and activation functions ($F_i$s and $G_i$s) have
to be stored on the microcontroller. Our paper presents an
approach for minimizing the memory usage of the implementation while satisfying
the hard constraints on the states.
%\section{Finite Precision Implementation}
%\Sadegh{To be included by Rocco}

### 2019-04-11 emsoft-2019/finite_precision.tex
comments and edits for section 3

% general blob about fixed-point arithmetic
The main task at compile time is to select the fixed-point format for each value
in the program. The format specifies the total word length and how many bits are
available for the integer part of the number. Given ranges on program inputs, a
range analysis estimates possibles ranges of all intermediate values in the
program, which in turn determine how many integer bits are sufficient to
avoid overflow. The remaining bits of the total word length represent the
fractional part of a number; more fractional bits provide more precision.

% introduces roundoff errors, these can be tracked with a dataflow analysis
Finite precision unavoidably introduces roundoff errors, which can accumulate
during the course of a computation.
A sound static roundoff error analysis computes a guaranteed upper bound on the
error of the result by tracking worst-case errors at every operation. Given a
word length for each value, several tools~\cite{Daisy,Gappa} automatically determine the number of
integer bits needed and compute roundoff errors using a dataflow analysis.
They track real-valued ranges at every intermediate operation. These ranges
determine the fixed-point formats and thus also the individual worst-case roundoff errors,
which the analyses track separately from the ranges. To ensure guaranteed
error bounds, both the ranges and errors are computed using interval~\cite{intervals} or
affine arithmetic~\cite{affineArithmetic} which compute sound enclosures. Since intervals do not
track correlations between variables and affine arithmetic tracks only linear
ones, the resulting ranges and errors are over-approximated.
% A standard technique to increase accuracy of the computed errors is interval
% subdivision, which divides the input domain into equally-sized subdomains and
% runs roundoff error analysis separately for each subdomain. The
% over-approximations on each subdomain tend to be smaller, which increases the
% overall tightness of range and error bounds.

% mixed-precision tuning
To reduce memory usage and increase efficiency, we want to choose as short word
lengths as possible. Mixed-precision tuning tools automatically determine possibly different
word lengths for different values. Compared to uniform precision (or word length),
mixed-precision often leads to improved resource usage, but due to the complexity of
fixed-point arithmetic and the error analysis, is challenging to do manually.  
Automated mixed-precision tuning tools usually perform a search: they repeatedly
select a candidate mixed-precision assignment (i.e. different word lengths for
different values) and check whether the assignment satisfies a given error
bound. The candidate assignments are chosen based on a heuristic which guides
the search towards promising candidates, e.g. using delta-debugging~\cite{DaisyTuning} or
simulated annealing~\cite{}. \eva{add missing refs}

>>>>

% general blob about fixed-point arithmetic
The main task at compile time is to select the fixed-point format for each input and intermediate value
in the program. The format specifies the total word length and how many bits are
available for the integer part of the number. Given ranges on program inputs, a
range analysis is usually used to estimate ranges of all intermediate values in the
program~\cite{Pang2011,Kinsman2009}, which in turn determine how many integer bits are sufficient to
avoid overflow. The remaining bits of the total word length represent the
fractional part of a number; more fractional bits provide more precision.

% introduces roundoff errors, these can be tracked with a dataflow analysis
Finite precision unavoidably introduces roundoff errors, which can accumulate
during the course of a computation.
A sound static roundoff error analysis computes a guaranteed upper bound on the
error of the result by tracking worst-case errors at every operation. Given a
word length for each value, several tools automatically determine the number of
integer bits needed and compute roundoff errors using a dataflow analysis~\cite{Daisy,Gappa}.
They track real-valued ranges at every intermediate operation. These ranges
determine the fixed-point formats and thus also the individual worst-case roundoff errors,
which the analyses track separately from the ranges. To ensure guaranteed
error bounds, both the ranges and errors are computed using interval~\cite{intervals} or
affine arithmetic~\cite{affineArithmetic} which compute sound enclosures. 
Affine arithmetic tracks linear correlations between variables, so for linear expressions
the computed ranges are as tight as possible (nonlinear arithmetic
leads to over-approximations).
% A standard technique to increase accuracy of the computed errors is interval
% subdivision, which divides the input domain into equally-sized subdomains and
% runs roundoff error analysis separately for each subdomain. The
% over-approximations on each subdomain tend to be smaller, which increases the
% overall tightness of range and error bounds.

% mixed-precision tuning
To reduce memory usage and increase efficiency, we want to choose as short word
lengths as possible. Mixed-precision tuning tools automatically determine possibly different
word lengths for different values~\cite{DaisyTuning,Lee2006}. Compared to uniform precision (or word length),
mixed-precision often leads to improved resource usage, but due to the complexity of
fixed-point arithmetic and the error analysis, is challenging to do manually.  
Automated mixed-precision tuning tools usually perform a search: they repeatedly
select a candidate mixed-precision assignment (i.e. different word lengths for
different values) and check whether the assignment satisfies a given error
bound. The candidate assignments are chosen based on a heuristic which guides
the search towards promising candidates, e.g. using delta-debugging~\cite{DaisyTuning} or
simulated annealing~\cite{Lee2006}.
